\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-29 paper.tex
improved academic writing style removed: 74 PASSIVE VOICE 7 TYPOGRAPHY ERRORS 34 WEASEL WORDS 7 STRUNK AND WHITE 6 ABBR (courtesy of https://github.com/devd/Academic-Writing-Check)

The internet of things (IoT) presents a great potential to positively influence
our daily work and life: domotics, assisted living, e-health and enhanced
learning are only a few examples\cite{atzori2010internet}. But with this
potential the IoT also presents a significant threat: by opening our smart
homes and personal data to all these interconnected devices, we also open them
to everyone who is able to break the virtual locks that protect them. IoT
devices typically have limited batteries, little processing power and memory is
restricted to the bare minimum. These properties make it hard to add any
additional service, including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally alerting the firewall if a
malicious pattern is detected\cite{denning1987intrusion}.

@@

Secondly, they don't support systematic reuse to create different
configurations for different devices, both dealing with hard- and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code in such a way that
redundant iterations, tests and computations are eliminated. Results show
substantially improved memory usage, execution time and usage of the wireless
radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
for patterns in the implementation of these classes of algorithms, details some
of the ideas behind FCF and identifies the different aspects of the problem
space. Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

These two topics lead to the first two classes: detecting intrusions and
cooperative decision making. We also identify software attestation as a third
class. Although it can be considered as a special case of a combination of both
other classes, its focus is not on the detection of intrusions, but on the
effect of an intrusion on the system itself.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
add the notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories offer very interesting prospects, the authors have to admit that the
impact on the resource-constrained devices might still be too much. We
therefore focus on the three categories agreed upon by most sources.

@@

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. Aberrant behavior is flagged as a
  possible intrusion, but this is no exact science. The baseline profile can
  evolve over time, especially in wireless and mobile environments. The flagged
  incidents should therefore also be handled with care and no absolute
  decisions can be made from single events.

  \item[Signature or misuse detection] starts with known patterns of attacks
  and tries to find evidence in the actions of and communication with a device.
  Because many attacks use valid ways to communicate with services on the
  devices, this can also produce false positives and offers no black and white
  certainty.

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although that this approach can detect
  previously unknown attacks, it can sometimes not detect specific attacks that
  can be detected by anomaly or misuse detection.

@@

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged intrusions. We identify
two ways by which this cooperative decision making can be constructed and focus
on the way participants exchange information: broadcasting and interactive.

@@

  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
  to be implemented locally. This is illustrated by the reputation-based
  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
  the same principle of a watchdog also found in \cite{mishra2004intrusion}.
  Using this algorithm, nodes can exchange information about the reputation of
  other nodes, solely with broadcasted messages. Combining this information
  allows them to decide about their trust in a given node, based on more than
  their own, often partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, a
  distributed decision can be made about the jointly identified intruder.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. It offers algorithms for exchanging
information about the software that is running on a device, and aims to
identify devices that can no longer prove that their content is unaltered. A
good example is SAKE\cite{seshadri2008sake}.

Software attestation is hard and many details can cause an algorithm to fail.
The general feeling is that even in the best conditions, there is always a way
to circumvent the most ingenious algorithm to perform software-based
attestation.

@@

Given this classification, we now introduce a design pattern that identifies
common aspects of these algorithms. It will allow us to define a strategy for
optimizing the combination of these algorithms. From the classification and
survey of several algorithms, we extract two activities common to most
algorithms: the inspection of network traffic and the evaluation of thresholds.

The former is trivial: network traffic is the carrier along which malicious
payloads reach the device or where patterns of communication can be observed
when other devices in the vicinity are affected by such payloads.

The latter is more complex: an algorithm based on signature or misuse detection
can maybe undeniably decide if an attack is real or not. But when dealing with
anomaly detection this line surely isn't that crisp. Especially in distributed
situations where even a neighboring device can not be trusted, detecting
intrusion is no absolute science. In this case one has to rely on statistics
and thresholds to decide when to drop trust in a given device.

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a devices
might miss the theoretical window in which certain actions are expected. Simply
reacting on such basic threshold might result in false positives. A way to
overcome this is to decouple the detection of problems and the decision what to
do about them.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

The different implementations would be linked and called in a sequential way,
when new messages are observed on the network and at a regular interval to
allow them to do internal housekeeping, such as threshold evaluation.

@@

  \item[Loops] by themselves are not a problem, but when different loops
  perform the same operations in sequence, an overhead might be incurred. An
  example from the ID algorithm pattern is the handling of messages: each
  algorithm parses each observed message on the network, processing the same
  byte stream, performing comparable operations to find similar patterns.

  \item[Common data] is cross-cutting the different algorithms, that have to
  store information about nodes within its own scope. This results in a lot of
  duplicated information in memory that also needs to be maintained.

  \item[The network] allows algorithms to operate on a distributed scale. But
  calls to the \ttt{SEND} function are scattered around the algorithms, causing
  the wireless radio to be accessed multiple times, keeping it active and
  energy consuming more than strictly needed.

@@

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost, e.g. in case a
new version of the algorithm is released, or simply when a different set of
algorithms is selected due to a changing risk analysis. Also, changing, or
simply implementing the algorithms from scratch, holds the risk of making
mistakes.

The heterogeneity of IoT devices results in many different software stacks,
each with e.g. their own network API. This also explains why there are very few
implementations readily available to developers: there is no common platform to
target for researchers. Even using a unified software library would not only
require a consistent implementation of it, but would also limit the operational
scope of the research implementation, still restricting it to a single language.

@@

To bridge the gap between the research world and the heterogeneous IoT where
developers want to implement an actual IDS, while still respecting the limited
resources they have at their disposal, we introduce \NAME, a framework
consisting of a DSL, code generator and software library.

The DSL enables researchers to formally describe their ID algorithms, without
focusing on any specific language or platform. Given such formal descriptions
of ID algorithms, a code generator is now able to combine these, while avoiding
the problems arising from the patterns/problems presented in listings
\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
different languages and platforms, while reusing a software library that
provides e.g. message parsing, network access wrapping,\dots

Choosing for a DSL and code generator is choosing a development strategy. This
implies several choices with respect to flexibility, update cost and
development effort. We will first position \NAME and motivate its design with
respect to these trade offs (\ref{positioning}). Second, we will look at the
principal design components of the DSL and how it deals with loops and common
data (\ref{dsl-design}). Third, the structure of the code generator will be
discussed (\ref{code-generator-design}). Finally, fourth, the supporting
software library is introduced (\ref{software-lib-design}).

@@

Figure \ref{fig:positioning} shows a classification of several development
strategies that can be adopted for projects involving wireless devices.

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

@@

Three properties are used to classify the strategies\cite{sadilek2007energy}:
flexibility, update cost and development effort.

@@

  \item[Flexibility] is a measure for the number of applications that can be
  deployed on a device without physical access.

@@

If we consider the trade off between flexibility and update cost, we can
distinguish two extremes: \emph{virtual machines} and \emph{full native
images}. Virtual machines allow for very small updates due to their concise
byte code, but at the same time are restricted by the limitations of their
instruction set and predefined capabilities. A full native image on the other
hand has unlimited flexibility, but requires transfer of a larger image over
the network.

Two intermediate alternatives are possible: \emph{modular native image} and
\emph{reconfigurable virtual machines}. With modular native images, only
modules are changed, limiting the update size, but the framework to support it
is static and limits the possibilities. Reconfigurable virtual machines take
this same approach, but apply it to virtual machines, allowing to
\emph{upgrade} the instruction set of the virtual machine.

Choosing a development strategy also determines the development cost: e.g.
developing for a virtual machine will be less demanding than targeting a native
image.

@@

Loops are a major contributor to the problem of algorithm fusibility. Loops in
program code are technical constructs and in fact often hide the actual
intended functionality.

To illustrate this, we take the example of parsing network messages. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional goal is to react when such pattern is present in a message. The loop
has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it. Using
event handling, constructions with loops are transformed to their functional
meaning, allowing identification and fusion of common functionalities.

The same goes for actively polling some variable until it reaches a certain
threshold. This also can be transformed to an event, eliminating the need for a
technical loop-construction.

@@

All algorithms deal with information about neighboring devices. To centralize
this information \NAME provides a unified \ttt{nodes} concept as part of its
DSL and exposes it as an object to the implementor.

Nodes can be extended with properties only available to the algorithm that
defines them. This way only one instance for every known device will be present
in memory. Finally, nodes can define a scope for some function to be executed
in. This way, nodes also offer a way to iterate all known devices, again
without the need for a loop.

@@

Access to the network that is scattered around an algorithm, causes the
wireless radio to be more activated than needed. Delaying the sending of
messages and grouping them, can lower this overhead.

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the nodes domain. An
underlying software framework will handle all interaction with the network,
triggering functions by means of events when new messages are parsed and
exposing an API to send messages through the nodes domain. This approach gives
the framework complete control over the way messages are sent, allowing
optimized marshaling, grouping,\dots

@@

To convert the DSL-based ID algorithms into production-ready code, we propose
the use of a code generator. Although many approaches here are applicable, like
template-based text expansion, XSLT transformations or even using the
CodeDom\cite{dollard2004code}, we believe that for the level of functional
transformations that need to be performed, a richer model is required.

A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations can now be applied to create a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit actual source code. Figure
\ref{fig:code-generation} shows the overall procedure.

@@

The DSL is parsed and imported into a semantic model. Based on the minimal type
information present in the model, all remaining types are inferred. Through a
series of semantic transformations, the model evolves from the functional to a
more technical one, a code model. This code model can be considered as a rich
abstract syntax tree (AST) with an extensive set of language constructs.
Syntactic transformations now lower this level of expressiveness according to
the desired platform and language. In the end the code model can be directly
emitted to the target language.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation. This way, the
functionality of different algorithms can be reorganized to allow for more
optimal execution and memory usage. Hence the name \NAME of the framework:
intrusion detection functionality organization optimization.

Figure \ref{fig:meta-model} shows the central part of the semantic model as
implemented for \NAME.

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
\label{fig:meta-model}
\end{figure}

The concept driving this model is the structure starting at the \emph{Model} up
to the \emph{Execution Strategy}. This structure encompasses the actual
\emph{Function}s and underlying \emph{Statement}s and allows for interpretation
and fusion of the algorithms.

@@

Thanks to the functional description of events and the effects that are
triggered by them, the code generator is able to identify the same actions
across different algorithms. This way it is possible to extract the parsing of
incoming messages, group actions on the same sets of data or scheduled function
executions.

FCF typically looks for patterns in the semantic model and performs model
transformations to change these patterns to a more optimized organization.
Semantic transformations are performed on the semantic model itself and end
with a transformation of the semantic model into a code model, which encodes
the semantics into technical code constructs.

@@

The code model can be compared to an AST for a very expressive language. It
contains constructs found in certain languages, but not in others. Therefore a
second round of model transformations needs to be performed. These syntactic
transformations morph the unsupported constructs into comparable constructs
that are supported by the targeted platform and language.

Depending on the target platform, some of the constructions used to e.g. parse
incoming messages, schedule tasks,\dots may not be transformed into
self-containing code. The generator then uses a software library to perform
these tasks. It will look for patterns that indicate any of these actions and
transform them into calls into the \NAME software library.

>>>>

The internet of things (IoT) presents a great potential to positively influence
our daily work and life through domotics, assisted living, e-health and
enhanced learning\cite{atzori2010internet}. But with this potential the IoT
also presents a significant threat: by opening our smart homes and personal
data to all these interconnected devices, we also open them to everyone who is
able to break the virtual locks that protect them. IoT devices typically have
limited batteries, processing power and memory. These properties make it hard
to add any additional service, including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally, after detecting a malicious
patter, alerting the firewall\cite{denning1987intrusion}.

@@

Secondly, they don't support systematic reuse to create different
configurations for different devices, both dealing with hard- and software.
Addressing this in a transparent and automated way is important for any kind of
development in this area.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show improved memory usage,
execution time and usage of the wireless radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
for patterns in the implementation of these classes of algorithms, details some
of the ideas behind FCF and identifies the different aspects of the problem
space. Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT\@. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

These two topics lead to the first two classes: detecting intrusions and
cooperative decision making. We also identify software attestation as a third
class. Although it is a combination of both other classes, its focus is not on
the detection of intrusions, but on the effect of an intrusion on the system
itself.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors agree with this topology and add the
notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. It tries to flag aberrant behavior as a
  possible intrusion, but this is no exact science. The baseline profile can
  evolve over time, especially in wireless and mobile environments. Single
  events should therefore not result in absolute decisions.

  \item[Signature or misuse detection] starts with known patterns of attacks
  and tries to find evidence in the actions of and communication with a device.
  Because attacks often use valid ways to communicate with services on the
  devices, this can also produce false positives and offers no black and white
  certainty.

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although that this approach can detect
  previously unknown attacks, it can sometimes not detect specific attacks that
  anomaly or misuse detection can.

@@

A group is often stronger than the sum of its individual components. Because
not all devices are always actively participating in the network, they can miss
clues that would otherwise lead them to detect intrusions. It is hardly
impossible for a single device to detect a complex attack by itself. Therefore
a second important research topic consists of cooperative algorithms to combine
information from single identities into a group-based decision about alleged
intrusions. We identify two ways to construct cooperative decision making and
focus on the way participants exchange information: broadcasting and
interactive.

@@

  \item[Broadcasting] can sometimes allow to implement a distributed,
  cooperative algorithm locally. The reputation-based detection algorithm
  introduced in \cite{ganeriwal2008reputation} illustrates this, based on the
  same principle of a watchdog also found in \cite{mishra2004intrusion}. Using
  this algorithm, nodes can exchange information about the reputation of other
  nodes, solely with broadcasts. Combining this information allows them to
  decide about their trust in a given node, based on more than their own, often
  partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. It offers algorithms for exchanging
information about the software that is running on a device and allows to
identify devices with modified content, that no longer can pass attestation. A
good example is SAKE\cite{seshadri2008sake}.

Software attestation is hard and details can cause an algorithm to fail. The
general feeling is that even in the best conditions, there is always a way to
circumvent the most ingenious algorithm to perform software-based attestation.

@@

Given this classification, we now introduce a design pattern that identifies
common aspects of these algorithms. It will allow us to define a strategy for
optimizing the combination of these algorithms. From the classification and a
survey of algorithms, we extract two activities common to most algorithms: the
inspection of network traffic and the evaluation of thresholds.

The former is trivial: network traffic is the carrier along which malicious
payloads reach a device directly. Indirectly a device can also detect aberrant
patterns of communication by other devices, as a result of a successful attack.

The latter is more complex: an algorithm based on signature or misuse detection
can maybe undeniably decide if an attack is real or not. But when dealing with
anomaly detection this line surely isn't that crisp.

Especially in distributed situations it is impossible to fully trust any
neighboring device. In this case one has to rely on statistics and thresholds
to decide when to drop trust in a given device.

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a devices
can miss a theoretical window in which one expects activity. Simply reacting on
such basic threshold can result in false positives. A way to overcome this is
to decouple the detection of problems and the decision what to do about them.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

The pattern calls the different, linked implementations in a sequential way in
two situations: first, when the device observes new messages on the network
and, secondly, at a regular interval to allow for internal housekeeping, such
as threshold evaluation.

@@

  \item[Loops] by themselves are not a problem, but different loops, who
  perform the same operations in sequence, can incur an overhead. An example
  from the ID algorithm pattern is the handling of messages: each algorithm
  parses each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

  \item[Common data] is cross-cutting the different algorithms, that have to
  store information about nodes within its own scope. This results in a lot of
  duplicated information in memory.

  \item[The network] allows algorithms to operate on a distributed scale. But
  scattered calls to the \ttt{SEND} function cause multiple accesses to the
  wireless radio, keeping it active and energy consuming more than strictly
  needed.

@@

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost. For example in
case of a release of a new version of the algorithm, or simply a different
selection of algorithms due to a changing risk analysis. Also, changing, or
simply implementing the algorithms from scratch, holds the risk of making
mistakes.

The heterogeneity of IoT devices results in different software stacks, each
with for example their own network API\@. This also explains why there are no
implementations readily available to developers: there is no common platform to
target for researchers. Even using a unified software library would not only
require a consistent implementation of it, but would also limit the operational
scope of the research implementation, still restricting it to a single language.

@@

To bridge the gap between the research world and the heterogeneous IoT where
developers want to implement an IDS, while still respecting the limited
resources they have at their disposal, we introduce \NAME, a framework
consisting of a DSL, code generator and software library.

The DSL enables researchers to formally describe their ID algorithms, without
focusing on any specific language or platform. Given such formal descriptions
of ID algorithms, a code generator is now able to combine these, while avoiding
the problems arising from the patterns/problems presented in listings
\ref{alg:id-algo-pattern} and \ref{alg:id-algo-application}. It can also target
different languages and platforms, while reusing a software library that
provides for example message parsing, network access wrapping,\dots

Choosing for a DSL and code generator is choosing a development strategy. This
implies choices with respect to flexibility, update cost and development
effort. We will first position \NAME and motivate its design with respect to
these trade offs (\ref{positioning}). Second, we will look at the principal
design components of the DSL and how it deals with loops and common data
(\ref{dsl-design}). Third, we discuss the structure of the code generator
(\ref{code-generator-design}) and finally, we introduce the supporting software
library (\ref{software-lib-design}).

@@

Figure \ref{fig:positioning} shows a classification of possible development
strategies for projects involving wireless devices.

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

@@

Three properties classify the strategies\cite{sadilek2007energy}: flexibility,
update cost and development effort.

@@

  \item[Flexibility] is a measure for the number of applications one can deploy
  on a device without physical access.

@@

If we consider the trade off between flexibility and update cost, we can
distinguish two extremes: \emph{virtual machines} and \emph{full native
images}. Virtual machines allow for small updates due to their concise byte
code, but their instruction set and predefined capabilities restrict them at
the same time. A full native image on the other hand has unlimited flexibility,
but requires transfer of a larger image over the network.

Two intermediate alternatives are possible: \emph{modular native image} and
\emph{reconfigurable virtual machines}. Modular native images allow changes on
module level, limiting the update size, but the framework to support it is
static and limits the possibilities. Reconfigurable virtual machines take this
same approach, but apply it to virtual machines, allowing to \emph{upgrade} the
instruction set of the virtual machine.

Choosing a development strategy also determines the development cost: for example
developing for a virtual machine will be less demanding than targeting a native
image.

@@

Loops are a major contributor to the problem of algorithm fusibility. Loops in
program code are technical constructs and in fact often hide the intended
functionality.

To illustrate this, we take the example of parsing network messages. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional goal is to react when such pattern is present in a message. The loop
has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it. Event
handling can transform constructions with loops to their functional meaning,
allowing identification and fusion of common functionality.

The same goes for actively polling some variable until it reaches a certain
threshold. When transformed to a reaction to an event, it eliminates the need
for a technical loop-construction.

@@

All algorithms deal with information about neighboring devices. To centralize
this information \NAME provides a unified \ttt{nodes} concept as part of its
DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way only one
instance for every known device will be present in memory. Finally, nodes can
define a scope to execute a function in. This way, nodes also offer a way to
iterate all known devices, again without the need for a loop.

@@

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them, can lower this overhead.

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the nodes domain. An
underlying software framework will handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the nodes domain. This approach gives the
framework complete control over the messages, allowing optimized marshaling,
grouping,\dots

@@

To convert the DSL-based ID algorithms into production-ready code, we propose
the use of a code generator. Although different approaches here are applicable,
like template-based text expansion, XSLT transformations or even using the
CodeDom\cite{dollard2004code}, we believe that the functional transformations
require a richer model.

A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations evolve the model into a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit source code. Figure \ref{fig:code-generation}
shows the overall procedure.

@@

After parsing the DSL, it's imported into a semantic model. The minimal type
information present in the initial model allows inferring of the remaining
types. Through a series of semantic transformations, the model evolves from the
functional to a more technical code model. This code model is comparable to a
rich abstract syntax tree (AST) with an extensive set of language constructs.
Syntactic transformations now lower this level of expressiveness according to
the desired platform and language. In the end the code model can be directly
emitted to the target language.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation. This way,
reorganizing the functionality of different algorithms allows for more optimal
execution and memory usage. Hence the name \NAME of the framework: intrusion
detection functionality organization optimization.

Figure \ref{fig:meta-model} shows the central part of the semantic model as
implemented for \NAME\@.

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \NAME\@. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
\label{fig:meta-model}
\end{figure}

The concept driving this model is the structure starting at the \emph{Model} up
to the \emph{Execution Strategy}. This structure encompasses the
\emph{Function}s and underlying \emph{Statement}s and allows for interpretation
and fusion of the algorithms.

@@

Thanks to the functional description of events and the effects triggered by
them, the code generator is able to identify the same actions across different
algorithms. This way it is possible to extract the parsing of incoming
messages, group actions on the same sets of data or scheduled function
executions.

FCF typically looks for patterns in the semantic model and performs model
transformations to change these patterns to a more optimized organization. It
first performs semantic transformations on the semantic model itself and ends
then transforms the semantic model into a code model, encoding the semantics
into technical code constructs.

@@

The code model can is comparable to an AST for an expressive language. It
contains constructs found in certain languages, but not in others. Therefore,
syntactic transformations morph the unsupported constructs into comparable
constructs supported by the targeted platform and language.

Depending on the target platform, some constructions aren't transformed into
self-containing code. The generator uses a software library for example in case
of parsing incoming messages and scheduling of tasks. It will look for patterns
that indicate any of these actions and transform them into calls into the \NAME
software library.
